Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-03-08 output/paper/manuscript.qmd
Started reviewing manuscript based on co-authors comments.

  val <- df[df$variable == exposure & df$outcome == outcome, ]$val
  val <- stringr::str_replace(
    string = val,
    pattern = "\\*",
    replacement = ""
  )
  glue::glue(
    "\\acr{{mc}} and $95\\%$ \\acr{{ci}}: {val}"
  )
}
```

@@

Epidemiological evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

The prevalence of several neurodevelopmental disorders has increased in the pediatric population [@GrandjeanLandrigan:2014], and multiple environmental pollutants may play a role in the increased rates of these disorders [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Multiple {{< acr edcs first_use=true >}}, ubiquitous chemicals present in many every-day products and diet, are capable of interfering with the endocrine system, and have shown associations with childhood neurodevelopment and behavior [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @TewarAuingerBraun:2016; @YuDuChiou:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @FurlongHerringBuckley:2017; @Braun:2017; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019; @ShoaffCoullWeuve:2020; @OhKimKannan:2023; @VilmandBeckBilenberg:2023]. Although both pregnancy and early childhood are crucial stages of (neuro)development, most of the available literature is focused on the effects of prenatal exposure to {{< acr edcs >}} on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

One group of {{< acr edcs >}} that may have a deleterious effect on neurodevelopment is the {{< acr opps >}}, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size and few {{< acr opp >}} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has been associated to several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to {{< acr bpa >}} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

@@

Neurodevelopmental outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the {{< acr ant first_use=true >}} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency in three different functions of attention: alerting, orienting, and executive attention. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found in @MaitreBontCasas:2018. The outcome of interest for the present study is the {{< acr hitrtse first_use=true >}} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high {{< acr hitrtse >}} indicates highly variable reactions, and is considered a measure of inattentiveness.

@@

Concentrations of the glucocorticosteroids were classified as quantifiable, below the {{< acr loq >}}, possible interference or out of range, and not detected. For each metabolite, we computed the fraction of values below the {{< acr loq >}} and not detected, both within each cohort and overall. We proceeded to impute these values using half the value of the corresponding {{< acr lloq >}}, for those metabolites that had less than `r threshold_within_lod`% of missings within each cohort and `r threshold_overall_lod`% of missings overall. Information about the {{< acr lloq >}} for the glucocorticosteroids is provided in @supptbl-lloq-mets. The remaining missing values were imputed using kNN from the `VIM` R package [@KowarikTempl:2016], for those metabolites that had less than `r threshold_within_nan`% of remaining missings within each cohort and `r threshold_overall_nan`% of remaining missings overall. We used `r knn_k` nearest neighbors. We natural log-transformed them to improve model fit, assessed with posterior predictive checks. To do so, replicated data were simulated with the fitted models and compared to the observed data. We used the `check_predictions` function from the `performance` R package using the default arguments [@LudeckeBen-ShacharPatil:2021]. Values of total cortisol, cortisone, and corticosterone production were expressed in nanograms per millilitre (ng/ml).

@@

We estimated {{< acr mcs >}} with the parametric g-formula, a method of standardization. The parametric g-formula involves the following steps: 1) fit a outcome model including both covariates and balancing weights; 2) create two new datasets identical to the original one but with the exposure shifted according to a user-specified intervention set by a deterministic function of the observed exposure levels; 3) use the outcome model to compute adjusted predictions in the two counterfactual datasets; 4) compute the difference between the means of the adjusted predictions in the counterfactual datasets. The causal parameter of interest $\phi$ was thus specified as the difference in the expected counterfactual outcomes under the shifted exposure levels $\left( \phi^\Delta = \mathbb{E} \left[ Y^{d_1} \right] - \mathbb{E} \left[ Y^{d_2} \right] \right)$. In order for this parameter to be identified, the usual causal identifiability conditions (no unmeasured confounding, positivity, and consistency) are required. Since these conditions are likely not satisfied, we focused on the estimation of a statistical estimand that is as close as possible to the causal parameter of interest.

@@

@tbl-pop-desc and @supptbl-pop-desc provide descriptive statistics for the outcome and covariates for the {{< acr helix >}} subcohort and for each cohort, respectively.
Of the 1,301 children of the {{< acr helix >}} subcohort, 1,297 had measurements of the non-persistent {{< acr edcs >}}. Measurements of the glucocorticosteroids were available for 1,004 children, of which 980 were matched to the {{< acr helix >}} subcohort. Measurements of both non-persistent {{< acr edcs >}} and glucocorticosteroids were available for 976 children of the subcohort. A flowchart describing the sample size for each research question is presented in @suppfig-flow-pop. The sample consisted of 55% girls. The median {{< acr hitrtse >}} was 300 ms ({{< acr iqr >}}, 231-368), with lower median values for EDEN, MOBA, and INMA, corresponding to the cohorts with older children. At the time of visit, the median age of the children was 8.06 years. The children were mostly Caucasian (90%), and the largest minority were of Pakistani origin (6.2%).

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific decrease in the levels of the exposures from the 90th to the 10th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}}. Most of the {{< acr cis first_use=true >}} included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`). The {{< acr opp >}} DETP was negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "HRT-SE")`).

@@

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for the {{< acr edcs >}} on the glucocorticosteroids and {{< acr hitrtse >}}. For {{< acr hitrtse >}}, significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "HRT-SE")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "HRT-SE")`) and MbZP (`r format_res(heffects1, "mbzp", "HRT-SE")`). For the glucocorticosteroids, significant differences were present across all three classes of {{< acr edcs >}} and for all outcomes. The largest differences were attributable to the {{< acr opps >}} DMTP (cortisol production, `r format_res(heffects2, "dmtp", "cortisol production")`) and DETP (corticosterone production, (`r format_res(heffects2, "detp", "corticosterone production")`); cortisone production, (`r format_res(heffects2, "detp", "cortisone production")`)). The forest plots of the individual {{< acr mcs >}} are presented in Supplementary Figures [-@suppfig-marginal-1sa] and [-@suppfig-marginal-2sa].

@Tbl-hypothesis-3 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for the glucocorticosteroids on {{< acr hitrtse >}}. Significant differences were present for cortisone production (`r format_res(heffects3, "cortisone prod.", "HRT-SE")`) and corticosterone production (`r format_res(heffects3, "corticost. prod.", "HRT-SE")`). Furthermore, for all exposures, the {{< acr mcs >}} had opposite sign (positive for males and negative for females). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-3sa.

@@

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} had negative effects on {{< acr hitrtse >}} and total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr cis >}} included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to {{< acr hitrtse >}}. The literature on {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several {{< acr edcs >}} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects are stronger in males. A major limitation of these studies is the reliance on spot urine samples, that might not be representative of long-term exposures.

Our results are consistent with prior epidemiological research that associated exposure to certain {{< acr edcs >}} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixture. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and {{< acr opps >}}, in relation to urinary glucocorticosteroids in childhood. Nonetheless, previous toxicological studies provide evidence for the inhibition by phthalates of human 11$\beta$-hydroxysteroid dehydrogenase 2 (11$\beta$-HSD2) activity, responsible for the conversion of active cortisol into inactive cortisone [@ZhaoChuHuang:2010; @MaLianDong:2011]. There is also *in silico* evidence suggesting that {{< acr bpa >}}, a phenol, and Triazophos (TAP), a organophosphorus insecticide, can bind to the human glucocorticoid receptor [@PrasanthDivyaSadasivan:2010; @YangLiRen:2019].

@@

Our findings should be interpreted in light of the following limitations and strengths. Limitations include the cross-sectional design of the present study. Importantly, the non-persistent {{< acr edcs >}} were measured in a pool of night and morning urine samples before the clinical visit, to represent exposure over the previous day, whereas the glucocorticosteroids were measured in the night urine sample. Although we included a wide range of confounders there is the possibility, as with other observational studies, of residual confounding, which might lead to a bias away from the null. Some of the confounders indicated in the adjustment sets had to be remove due to large fractions of missing values. There is further the possibility of misspecification of the outcome model, although we included a spline of the exposure to relax some of the linearity assumptions. The use of more data-adaptive learners was excluded due to the relatively small sample size. We finally acknowledge the possibility that some of chemicals might not act independently (mixture effect). Further research is thus warranted.

Strengths of the present study include the use of pooled urine samples for chemical assessment, since it is known that these specific {{< acr edcs >}} have very short half-lives [@CasasBasaganaSakhi:2018].
We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to try to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of the regression models, but by making use of the g-computation procedure and estimate {{< acr mcs >}}.

In conclusion, in a study of 1,297 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent {{< acr edcs >}} in childhood might have short-term effects on {{< acr hitrtse >}} in childhood, (ii) exposure to non-persistent {{< acr edcs >}} in childhood might disrupt the {{< acr hpa >}} axis in childhood, and (iii) disruption of the {{< acr hpa >}} axis in childhood might have short-term, sex-specific effects on {{< acr hitrtse >}}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to {{< acr edcs >}} on childhood neurodevelopment in larger populations.

>>>>

  val <- df[df$variable == exposure & df$outcome == outcome, ]$val
  val <- stringr::str_replace(
    string = val,
    pattern = "\\*",
    replacement = ""
  )
  val1 <- stringr::str_split(val, " ", n = 2)[[1]][1]
  val2 <- stringr::str_split(val, " ", n = 2)[[1]][2]
  glue::glue(
    "MC: {val1} and $95\\%$ \\acr{{ci}}: {val2}"
  )
}
```

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

The prevalence of several neurodevelopmental disorders has increased in the pediatric population [@GrandjeanLandrigan:2014], and multiple environmental pollutants may play a role in the increased rates of these disorders [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Multiple {{< acr edcs first_use=true >}}, ubiquitous chemicals present in many every-day products and diet, are capable of interfering with the endocrine system, and have shown associations with childhood neurodevelopment and behavior [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @TewarAuingerBraun:2016; @YuDuChiou:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @FurlongHerringBuckley:2017; @Braun:2017; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019; @ShoaffCoullWeuve:2020; @OhKimKannan:2023; @VilmandBeckBilenberg:2023]. Although both pregnancy and early infancy are crucial stages of (neuro)development, most of the available literature is focused on the effects of prenatal exposure to {{< acr edcs >}} on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

One group of {{< acr edcs >}} that may have a deleterious effect on neurodevelopment is the {{< acr opps >}}, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has been associated with several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to {{< acr bpa >}} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

@@

Neurodevelopmental outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the {{< acr ant first_use=true >}} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency in three different components of attention: alerting, orienting, and executive attention. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found in @MaitreBontCasas:2018. The outcome of interest for the present study is the {{< acr hitrtse first_use=true >}} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high {{< acr hitrtse >}} indicates highly variable reaction times, and is considered a measure of inattentiveness.

@@

Concentrations of the glucocorticosteroids were classified as quantifiable, below the {{< acr loq >}}, possible interference or out of range, and not detected. For each metabolite, we computed the fraction of values below the {{< acr loq >}} and not detected, both within each cohort and overall. We proceeded to impute these values using half the value of the corresponding {{< acr loq >}}, for those metabolites that had less than `r threshold_within_lod`% of missings within each cohort and `r threshold_overall_lod`% of missings overall. Information about the {{< acr lloq >}} for the glucocorticosteroids is provided in @supptbl-lloq-mets. The remaining missing values were imputed using kNN from the `VIM` R package [@KowarikTempl:2016], for those metabolites that had less than `r threshold_within_nan`% of remaining missings within each cohort and `r threshold_overall_nan`% of remaining missings overall. We used `r knn_k` nearest neighbors. We natural log-transformed them to improve model fit, assessed with posterior predictive checks. To do so, replicated data were simulated with the fitted models and compared to the observed data. We used the `check_predictions` function from the `performance` R package using the default arguments [@LudeckeBen-ShacharPatil:2021]. Values of total cortisol, cortisone, and corticosterone production were expressed in nanograms per millilitre (ng/ml).

@@

We estimated {{< acr mcs >}} with the parametric g-formula, a method of standardization. The parametric g-formula involves the following steps: 1) fit a outcome model including both covariates and balancing weights; 2) create two new datasets identical to the original one but with the exposure shifted according to a user-specified intervention set by a deterministic function of the observed exposure levels; 3) use the outcome model to compute adjusted predictions in the two counterfactual datasets; 4) compute the difference between the means of the adjusted predictions in the counterfactual datasets. The causal parameter of interest was thus specified as the difference in the expected counterfactual outcomes under the shifted exposure levels $\left( \mathbb{E} \left[ Y^{d_1} \right] - \mathbb{E} \left[ Y^{d_2} \right] \right)$. In order for this parameter to be identified, the usual causal identifiability conditions (no unmeasured confounding, positivity, and consistency) are required. Since these conditions are likely not satisfied, we focused on the estimation of a statistical estimand that is as close as possible to the causal parameter of interest.

@@

@tbl-pop-desc and @supptbl-pop-desc provide descriptive statistics for the outcome and covariates for the {{< acr helix >}} subcohort and for each cohort, respectively.
Of the 1,301 children of the {{< acr helix >}} subcohort, 1,297 had measurements of the non-persistent {{< acr edcs >}}. Measurements of the glucocorticosteroids were available for 1,004 children, of which 980 were matched to the {{< acr helix >}} subcohort. Measurements of both non-persistent {{< acr edcs >}} and glucocorticosteroids were available for 976 children of the subcohort. A flowchart describing the sample size for each research question is presented in @suppfig-flow-pop. The sample consisted of 55% males. The median {{< acr hitrtse >}} was 300 ms ({{< acr iqr >}}, 231-368), with lower median values for EDEN, MOBA, and INMA, corresponding to the cohorts with older children. At the time of visit, the median age of the children was 8.06 years. The children were mostly Caucasian (90%), and the largest minority were of Pakistani origin (6.2%).

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific decrease in the levels of the exposures from the 90th to the 10th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}} and thus lower attention. Most of the {{< acr cis first_use=true >}} included the null effect, though. Significant effects were observed for the paraben MEPA (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`). The {{< acr opp >}} DETP was negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "HRT-SE")`).

@@

@Tbl-hypothesis-1and2 presents the results of the difference between estimates of the {{< acr mcs >}} for females and males, for the {{< acr edcs >}} on the glucocorticosteroids and {{< acr hitrtse >}}. For {{< acr hitrtse >}}, significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "HRT-SE")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "HRT-SE")`) and MbZP (`r format_res(heffects1, "mbzp", "HRT-SE")`). For the glucocorticosteroids, significant differences were present across all three classes of {{< acr edcs >}} and for all outcomes. The largest differences were attributable to the {{< acr opps >}} DMTP (cortisol production, `r format_res(heffects2, "dmtp", "cortisol production")`) and DETP (corticosterone production, (`r format_res(heffects2, "detp", "corticosterone production")`); cortisone production, (`r format_res(heffects2, "detp", "cortisone production")`)). The forest plots of the individual {{< acr mcs >}} are presented in Supplementary Figures [-@suppfig-marginal-1sa] and [-@suppfig-marginal-2sa].

@Tbl-hypothesis-3 presents the results of the difference between estimates of the {{< acr mcs >}} for females and males, for the glucocorticosteroids on {{< acr hitrtse >}}. Significant differences were present for cortisone production (`r format_res(heffects3, "cortisone prod.", "HRT-SE")`) and corticosterone production (`r format_res(heffects3, "corticost. prod.", "HRT-SE")`). Furthermore, for all exposures, the {{< acr mcs >}} had opposite sign (positive for males and negative for females). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-3sa.

@@

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} was associated with {{< acr hitrtse >}}, a measure of inattentiveness, and with total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr cis >}} included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to {{< acr hitrtse >}}. The literature on non-persistent {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of non-persistent {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].

Our results are consistent with prior epidemiological research that associated exposure to certain non-persistent {{< acr edcs >}} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixture. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and {{< acr opps >}}, in relation to urinary glucocorticosteroids in childhood. Nonetheless, previous toxicological studies provide evidence for the inhibition by phthalates of human 11$\beta$-hydroxysteroid dehydrogenase 2 (11$\beta$-HSD2) activity, responsible for the conversion of active cortisol into inactive cortisone [@ZhaoChuHuang:2010; @MaLianDong:2011]. There is also *in silico* evidence suggesting that {{< acr bpa >}}, a phenol, and Triazophos (TAP), a organophosphorus insecticide, can bind to the human glucocorticoid receptor [@PrasanthDivyaSadasivan:2010; @YangLiRen:2019].

@@

Our findings should be interpreted in light of the following limitations and strengths. Limitations include the cross-sectional design of the present study. Importantly, the non-persistent {{< acr edcs >}} were measured in a pool of night and morning urine samples before the clinical visit, to represent exposure over the previous day, whereas the glucocorticosteroids were measured in the night urine sample. Although we included a wide range of confounders there is the possibility, as with other observational studies, of residual confounding, which might lead to a bias away from the null. Some of the confounders indicated in the adjustment sets had to be removed due to large fractions of missing values. There is further the possibility of misspecification of the outcome model, although we included a spline of the exposure to relax some of the linearity assumptions. The use of more data-adaptive learners was excluded due to the relatively small sample size. We finally acknowledge the possibility that some of chemicals might not act independently (mixture effect). Further research is thus warranted.

Strengths of the present study include the use of pooled urine samples for chemical assessment to obtain more representative long-term exposures, since it is known that these specific {{< acr edcs >}} have very short half-lives [@PerrierGiorgis-AllemandSlama:2016; @CasasBasaganaSakhi:2018]. We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to try to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of the regression models, but by making use of the g-computation procedure and estimate {{< acr mcs >}}.

In conclusion, in a study of 1,297 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent {{< acr edcs >}} in childhood might have short-term effects on {{< acr hitrtse >}} in childhood, (ii) exposure to non-persistent {{< acr edcs >}} in childhood might disrupt the {{< acr hpa >}} axis in childhood, and (iii) disruption of the {{< acr hpa >}} axis in childhood might have short-term, sex-specific effects on {{< acr hitrtse >}}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to non-persistent {{< acr edcs >}} on childhood neurodevelopment in larger populations.
